---
title: "Advanced Machine Learning (732A96) - Lab 02"
author: "Hector Plata (hecpl268)"
output: 
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of the lab is to put in practice some of the concepts convered in the elctures. To do so, you are asked to model the behaviour of a robot that walks around a ring. The ring is divided into 10 sectors. At any give time point, the robot is in one of the sectors and decides with equal probability to stay in that sector or move to the next sector. You do not have direct observation of the robot. However, the robot is equipped with a tracking device that you can access. The device is not very accurate thought: If the robot is in the sector $i$, then th edevice will report that the robot is in the sectors $[i-2, i+2]$ with equal probability.

# Task 1

Build a hidden Markon Model (HMM) for the scenario described above.

```{r}
# Defining the States and Symbols.
states = paste(rep("Z", 10), 1:10, sep="")
symbols = paste(rep("X", 10), 1:10, sep="")

# Transition probability matrix.
P = diag(0, 10, 10)
P[1:9, 2:10] = diag(0.5, 9, 9)
P = P + diag(0.5, 10, 10)

# Emision probability matrix.
E = matrix(0, 10, 10)

for(i in 1:10)
{
  for(j in c(-2,-1,0,1,2))
  {
    if (i + j < 1)
    {
      k = 10 + i + j
    }
    else if (i + j > 10)
    {
      k = i - 10 + j
    }
    else
    {
      k = i + j
    }
    print(i)
    print(k)
    E[i, k] = 1/5
  }
}

# Defining the HMM.
hmm = initHMM(States=states,
              Symbols=symbols,
              transProbs=P,
              emissionProbs=E)

print(hmm)
```

# Task 2

Simulate the HMM for 100 time steps.

```{r}
set.seed(4)
sim = simHMM(hmm, 100)
print(sim)
```

# Task 3

Discard the hidden states from the sample obtained above. Use the remaining observations to compute the filtered and smoothed probability distributions for each of the 100 time points. Compute also the most probable path.


```{r}
sim_states = sim$states
sim_obs = sim$observation
filtered = prop.table(exp(forward(hmm, sim_obs)), 2)
smooth = posterior(hmm, sim_obs)
most_prob_path = viterbi(hmm, sim_obs)
print(most_prob_path)
```

# Task 4

Compute the accuracy of the filtered and smoothed probability distributions, and of the most probable path. That is, compute the percentage of the true hidden states that are guessed by each method.

**Hint:** Note that the function `forward`in the `HMM`package returns probabilities in log scale. You may need to use the functions `exp` and `prop.table` in order to obtain a normalized probability distribution. You may also want to use the functions `apply` and `which.max`to find out the most probable states. Finally, recall that you canc ompare two vector A and B elementwise as A==B, and that the function `table`will count the number of times that the different elements in a vector occur in the vector.

```{r}

```

```{r}

```